---
title: "Chapter 1 - Displaying Data"
output: html_document
---

*Notes from working through [An Introduction to the Science of Statistics: From Theory to Implementation, Preliminary Edition](https://www.math.arizona.edu/~jwatkins/statbook.pdf) by Joseph C. Watkins*

*With additional material from OpenSTAX [Introductory Statistics](https://openstax.org/details/books/introductory-statistics) book, by Barbara Illowsky and Susan Dean.*

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

## Types of Data

* Qualitative (Categorical) data, aka levels or factors
* Quantitative (Numerical) data
  * Discrete (counting)
  * Continuous

### Levels of Measurement

**Nominal scale** - Unordered categorical data.  Example, favorite foods.  Sushi is not inherently less than or greater than pizza.

**Ordinal scale** - Ordered categorical data, but difference between values cannot be measured.  Example, "strongly agree", "agree", "disagree", "strongly disagree".  (Qualitative/categorical?)

**Interval scale** - Ordered quantitative data where differences can be measured, but ratios cannot be taken because there is no true 0 point.  Example, Fahrenheit scale, where the difference between 60 and 80 is 20, but where 80 is not four times as hot as 20.

**Ratio scale** - Ordered quantitative data where differences and ratios can both be calculated.  Example, Kelvin scale or test scores counted from 0.

### Categorical Graphics

Simple Pie Chart
```{r}
males <- c(58, 18, 16, 7, 1)
pie(males)
```

Fancier Pie Chart (*note: backslash before % in original*)
```{r}
colors <- c("white", "grey70", "grey90", "grey50", "black")
male_labels <- round(males/sum(males) * 100, 1)
male_labels <- paste(male_labels, "%", sep=" ")
pie(males,
    main = "Proportion of US Male AIDS Cases by Transmission",
    col = colors,
    labels = male_labels,
    cex = 0.8
)
legend("topright",
       c("Male-male contact",
         "Injection drug use (IDU)",
         "High-risk heterosexual contact",
         "Male-male contact and IDU",
         "Other"),
       cex = 0.8,
       fill = colors
       )

```

Simple bar chart
```{r}
barplot(males)
```

Fancier bar plot, with females included this time.
```{r}
females <- c(0, 71, 27, 0, 2)
hiv <- array(c(males, females), dim=c(5,2))
barplot(hiv,
        main = "Proportion of US AIDS Cases by Sex and Transmission",
        ylab = "percent",
        beside = TRUE,
        names.arg = c("Males", "Females"),
        col = colors
)
legend("topright",
       c("Male-male contact",
         "Injection drug use (IDU)",
         "High-risk heterosexual contact",
         "Male-male contact and IDU",
         "Other"),
       cex = 0.8,
       fill = colors
)
```

Contingency tables, also known as two-way tables, cross tabulation tables, or cross classifying tables.

Let's get some smoking data to work with.
```{r}
smoking <- matrix(c(400, 1380, 416, 1823, 188, 1168), ncol = 3)
colnames(smoking) <- c("2 parents", "1 parent", "0 parents")
rownames(smoking) <- c("smokes", "does not smoke")
smoking
```

Now let's show the parts of a contingency table, transposed (I'm not sure how to show it all yet.)  With totals showing the *marginal distributions*, it should look something like this according to the book.  [I'm not sure how to add labels to the total column and rows.  Also, there might be an easier way to do this.]
```{r}
s <- cbind(t(smoking), rowSums(t(smoking)))
s <- rbind(s, cbind(t(colSums(t(smoking))), sum(smoking)))
s
```

Or, more nicely formatted, like
```{r}
knitr::kable(s)
```

Here is the same data as a segmented bar graph.
```{r}
barplot(smoking, legend = rownames(smoking))
```

### Histograms

Skewness is the direction of the longer tail.

Sturges' function for number of histogram bins: $\lfloor1+\log_2 n\rfloor$.

President ages at inauguration
```{r}
age <- c(57, 61, 57, 57, 58, 57, 61, 54, 68, 51, 49, 64, 50, 48, 65, 52, 56, 46, 54, 49, 51, 47, 55, 55, 54, 42, 51, 56, 55, 51, 54, 51, 60, 61, 43, 55, 56, 61, 52, 69, 64, 46, 54, 47, 70)
age
ecdf(age)
```

```{r}
par(mfrow = c(1, 2))
hist(age)
plot(ecdf(age), xlab = "age", main = "Age of Presidents at the Time of Inauguration", sub = "Empirical Cumulative Distribution Function")

```

**Empirical cumulative distribution function**:
$$ F_n(x) = \frac{1}{n} \#\text{(observations <= x)} $$

Alternate way to graph ecdf by sorting entries
```{r}
plot(sort(age),
     1:length(age)/length(age),
     type = "s",
     xlab = "age",
     ylab = "cumulative fraction",
     ylim = c(0, 1)
)
grid()
```

*Exercise 1.14* - Fraction of presidents under 60 at inauguration.
```{r}
sum(age < 60, na.rm = T) / length(age)
```

**Survivor function**  (*note: text says >=, not >*)
$$ S_n(x) = \frac{1}{n} \# \text{(observations > x)} = 1 - F_n(x) $$

### Scatterplot

Two-dimensional data
```{r}
femur <- c(38, 56, 59, 64, 74)
humerous <- c(41, 63, 70, 72, 84)
plot(femur, humerous, main = "Bone lengths for Archeopteryx")
```

