---
title: "Chapter 5 - Basics of Probability"
output: html_document
---

*Notes from working through [An Introduction to the Science of Statistics: From Theory to Implementation, Preliminary Edition](https://www.math.arizona.edu/~jwatkins/statbook.pdf) by Joseph C. Watkins*

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

## 5.1 Introduction

A *probability model* has

* $\Omega$ - The **sample space**, the set of possible outcomes.
* $A$ - An **event** - A collection of outcomes
$$
\begin{align*}
A &= \{\omega_1, \omega_2, \cdots, \omega_n\} \\
A &= \{\omega ; \omega \text{ has property } \mathcal P\}
\end{align*}
$$
* $P$ - **Probability** - A function that assigns a number to each event.

**Probability** - A branch of mathematics based on 3 axioms originally inspired by games of chance.

## 5.2 Axioms of Probability

1. $P(A) \geq 0$
2. $P(\Omega) = 1$
3. If two events are **mutually exclusive** or **disjoint** (i.e. $A \cap B = \emptyset$), then $P(A \cup B) = P(A) + P(B)$
*  The calculus version of axiom 3, still for mutually exclusive events...
$$ P\left( \, \bigcup_{j=1}^\infty A_j \right) = \sum_{j=1}^\infty P(A_j)$$

## 5.3 Consequences of the Axioms

1. The **complement** of $A$ is $A^c = \{\omega;\omega \notin A\}$
$$ P(A^c) = 1 - P(A) $$
$$ P(A) + P(A^c) = P(A \cup A^c) = P(\Omega) = 1 $$
2. The **difference** $B \setminus A$ are outcomes in $B$ but *not* in $A$.
$$ \text{if } A \subset B , ~~~ P(B \setminus A) = P(B) - P(A) $$
3. **Monotonicity** Symbol $\subset$ means **contains in**
$$ \text{if } A \subset B, ~~~ P(A) \leq P(B) $$
$$ P(A) \leq P(\Omega) = 1 $$
That, combined with axiom 1, means probability is always between 0 and 1 (inclusive)
$$ 0 \leq P(A) \leq 1 $$
4. **Inclusion-Exclusion**
$$ P(A \cup B) = P(A) + P(B) - P(A \cap B) $$

5. **The Bonferroni Inequality**
$$ P(A \cup B) \leq P(A) + P(B) $$

6. **Continuity Property** [See book, basically nested stuff acts nested?]

## 5.4 Counting

“In the case of equally likely outcomes, ﬁnding the probability of an event A is the result of two counting problems - namely finding $\#(A)$, the number of outcomes in the event A and finding $\#(\Omega)$, the number of outcomes in the sample space, $\Omega$.”

### 5.4.1 Fundamental Principle of Counting
The fundamental principle of counting says if you have two experiments, with $n_1$ and $n_2$ possible outcomes respectively, then in total there are $n_1 \times n_2$ possible outcomes.

An example would be the number of ways of choosing a president and vice-president from a group of $n$ people. Assuming one person could not hold both posts, the number of outcomes is $n \times (n−1)$

*Exercise 5.10* Find the # of ways to draw two cards? $52 \times 52$ if you reshuffle in between? $52 \times 51$ if you do not reshuffle. If you do not reshuffle, the number of ways to draw two aces is $4 \times 3$.

*Exercise 5.11* Generalizing the **fundamental principle of counting** to $k$ experiments, I think you get...
$$ \text{number of outcomes} = \sum_{i=1}^k n_i $$

### 5.4.2 Permutations
**Permutation** - When you have $n$ things, and pick out an **ordered** subset of $k$ of those things.
$$ \text{permutation outcomes } = n \times (n - 1) \times \cdots \times (n - k + 1) = (n)_k = \frac{n!}{(n-k)!} $$

Note that $(n)_k$ is pronounced "n falling k".

*Example 5.12* - Birthday problem, the probability among $k$ people that at least two people share a birthday.</p>
$$ P\{\text{at least one pair share birthdays}\} = 1 - \frac{(365)_k}{365^k} $$
```{r}
bday <- data.frame(people = c(1:100), prob = rep(1, 100))
for (k in 2:100) {bday$prob[k] = bday$prob[k-1]*(365-k+1)/365}
bday$compl = 1 - bday$prob
plot(bday$people, bday$compl)
```

### 5.4.3 Combinations
**Combination** - When you have $n$ things, and pick out a subset of $k$ of those things such that **order does not matter**.
$$ \text{combination outcomes } = \frac{(n)_k}{k!} = {n \choose k} = \frac{n!}{k!(n-k)!} $$

Note that ${n \choose k}$ is pronounced "n choose k".

*Exercise 5.17* - Verify the identities 
$$ {n \choose 1} = {n \choose n-1} = n ~~~ \text{and} ~~~ {n \choose k} = {n \choose n-k} $$

```{r}
choose(8, 3)
```


