---
title: "Chapter 7 - Random Variables and Distribution Functions"
output: html_document
---

*Notes from working through [An Introduction to the Science of Statistics: From Theory to Implementation, Preliminary Edition](https://www.math.arizona.edu/~jwatkins/statbook.pdf) by Joseph C. Watkins*

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

## 7.1 Introduction

| Statistics | Probability |
| --- | --- |
| Empirical cumulative distribution function | Cumulative distribution function |
| Sample means and variances | Distributional means and variances |

**Random Variable** - A function that takes in something from the probability space and spits out a real number.
$$ X:\Omega \rightarrow \mathbb{R} $$

## 7.2 Distribution Functions

A **cumulative distribution function** of a random variable X is defined by
$$ F_X(x) = P\{\omega \in \Omega ; X(\omega) \leq x\} $$
Or, in more abbreviated notation...
$$ F_X(x) = P\{X \leq x\} $$
*Exercise 7.9* Let $X$ be the number of heads on three independent flips of a *biased* coin that gives heads with probability $p$.  Give the cumulative distribution function $F_X$ for $X$.

*Exercise 7.10* Let $X$ be the number of spades in a collection of three cards.  Give the cumulative distribution function for $X$ and plot it with R.

| Spades | 0 | 1 | 2 | 3 |
| --- | --- | --- | --- | --- |


```{r}
spades <- 0:3
f <- ( (choose(13, spades) * choose(39, 3 - spades)) / choose(52,3) )
f
Fx <- cumsum(f)
Fx
plot(spades, Fx, type="s", ylim=c(0,1))
```

*Exercise 7.11* Find the cumulative distribution function of $Y=X^3$ in terms of $F_X$, the distribution function for $X$.

## 7.3 Properties of Distribution Function

A distribution function $F_X$ starts at 0, ends at 1, and does not decrease as $x$ increases.

*Exercise 7.17* [See book] $\pi \left(\frac23\right)^2 - \pi \left(\frac13\right)^2 = \frac13 \pi$

$$
\begin{align*}
A &= \pi r^2 & A&=\pi 1^2 & A=\pi   \\
r &= \sqrt\frac{A}{\pi} \\
\text{3rd quartile, } r_{third} &= \sqrt \frac34 \\
\text{median, } r_{median} &= \sqrt \frac12 \\
\text{1st quartile, } r_{first} &= \sqrt \frac14 \\
\end{align*}
$$

I believe distribution functions in R are prefixed with `p` for probability (such as `pnorm()`).



## 7.4 Mass Functions

The **(probability) mass function** of a discrete random variable $X$ is
$$ f_X(x) = P\{X = x\} $$

It is always positive, and sums to 1 for all $x$.


## 7.5 Density Functions


## Dean's Distribution Interlude

Here are some distributions graphed in R, with the mass or density function in black, and the cumulative distribution function in blue.

The normal distribution.
```{r}
curve(pnorm, xlim=c(-3,3), col="blue") # Normal cumulative probability.
curve(dnorm, add=T)                    # Normal density
abline(h=.5, v=0, col="lightgray")
```
```

The binomial distribution (for 50 trials, 50% probability)
```{r}
plot(pbinom(q=0:50, size=50, prob=0.5), type="s", col="blue")
```

```{r}
normies <- rnorm(100000)
plot(cumsum(sort(normies)) / 100000)
hist(normies, freq=TRUE, breaks=seq(-5,5,.1))

```





