---
title: "Triola"
output: html_document
---

Notes from working through *Essentials of Statistics, 6th Ed.*, by Mario F. Triola

```{r setup, comment=NA}
knitr::opts_chunk$set(comment = NA)
```

## Chapter 4 - Probability

### Chapter 4-2 - Addition Rule and Multiplication Rule

1. $P(A)$ is the probability that a randomly selected adult has blue eyes
   $P(\bar A)$ is the probability that a randomly selected adult does not have blue eyes.
   
3. They are dependent, but can be considered independent for the purposes of calculations because the sample is (well) under 5% of the population.

5. 74%

7. $P(\bar N) = .670$, and represents surveyed adultswho do travel on commercial flights.

9. 67.6%
```{r}
1 - ((329 + 33) / (329 + 264 + 249 + 145 + 33 + 54 + 31 + 13))
```

11. 906. No, they are not disjoint.
```{r}
qsr <- matrix(c(329, 33, 264, 54, 249, 31, 145, 13), nrow = 2)
rownames(qsr) <- c("Order Accurate", "Order Inaccurate")
colnames(qsr) <- c("McDonald's", "Burger King", "Wendy's", "Taco Bell")
qsr
(sum(qsr[1,]) + sum(qsr[,1]) - qsr[1,1]) / sum(qsr)
```

13. a) Yes, the events are independent, and the probability is .01997
    b) No, the events are dependent, and the probability is 0.01986 

```{r}
(sum(qsr[, "Taco Bell"]) / sum(qsr))^2
sum(qsr[, "Taco Bell"]) / sum(qsr) * (sum(qsr[, "Taco Bell"]) - 1) / (sum(qsr) - 1)
```

15. a) Yes, the events are independent, and the probability is .7794
    b) No, the events are dependent, and the probability is .7793
```{r}
(sum(qsr["Order Accurate",]) / sum(qsr))^2
sum(qsr["Order Accurate",]) / sum(qsr) * (sum(qsr["Order Accurate",]) - 1) / (sum(qsr) - 1)
```

19. .0156

```{r}
sum(qsr[,"Wendy's"]) / sum(qsr) * (sum(qsr[,"Wendy's"])-1) / (sum(qsr) - 1) * (sum(qsr[,"Wendy's"])-2) / (sum(qsr) - 2) 

(factorial(sum(qsr[,"Wendy's"])) / factorial(sum(qsr[,"Wendy's"]) - 3)) / (factorial(sum(qsr)) / factorial(sum(qsr) - 3))

```

### 4.3 Complements, Conditional Probability, and Bayes' Theorem

5. Probability of 3 boys = $\frac12^3 = \frac18$.  Probability of the complement, at least one girl, = $\frac78$, or .875

7. .982

```{r}
1 - .512^6
```

9. .344

```{r}
1 - .9^4
```

11. .965

```{r}
1 - .8^15
```

13. a. .628
    b. .372
    c. That if you give someone the right denomination to purchase something, they are more likely to do so compared to breaking a higher denomination.
```{r}
27 / (27 + 16)
16 / (27 + 16)
```

17. .00173

```{r}
2/1155
335/337
```

19. .994

21. a. .9991
    b. .999973

```{r}
1 - .03^2
1 - .03^3
```

23. .490

```{r}
1 - (1-.126)^5
```

25. Birthday, 25 people, .569

```{r}
1 - prod((365 - 0:24) / 365)
```

### 4.4 Counting

3. Permutations rule with repeated values?, because order matters
   NOPE: Multiplication rule.
   
5. 1/10000, or .0001

7. 684, nope 1/171

```{r}
factorial(19) / (factorial(17) * factorial(2))
```

9. 1/40320, or .0000248

```{r}
1 / factorial(8)
```

11. 1/254251200

```{r}
factorial(50) / factorial(50 - 5)
```

13. 1/100,000,000.  No

```{r}
100^4
```

15. 1/1820

```{r}
choose(16,4)
factorial(16) / (factorial(16-4) * factorial(4))
```

17. 1/292,201,338

```{r}
choose(69, 5) * 26
```

19. 1/100,000

21. 800, 6,400,000,000

```{r}
800 * 800 * 10000
```

23. a. 5040
    b. 210
    c. 1/5040, nope 1/210

```{r}
factorial(10)/factorial(10-4)
factorial(10)/(factorial(10-4)*factorial(4))
```

25. 40,320, 1/40,320

```{r}
factorial(8)
```

27. 20,922,790,000,000, no 653,837,184,000, forgot repeat letters.


```{r}
factorial(16) / (2*2*2*2*2)
```

31. 62

```{r}
2^5 + 2^4 + 2^3 + 2^2 + 2^1
```

33. 4.7%.  Nope, 4.8% 

```{r}
(4/52) * (16/52) * 2
```

35. 

